The Castro AMR Simulation Code: Current and Future Developments

### 2020-02-04 astronum_2019_Castro/astronum-castro-2019.tex
update
Fixed typo in Castro astronum proceedings.

In the SDC formalism~\cite{dutt:2000,minion:2003}, we start by writing our system in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc} + \Rbs{\Uc} \right ] dt
\end{equation}
We next introduce the idea of iteratively approximating the solution.
We'll use the superscript $k$ to denote the current iteration and
$k+1$ for the new iteration.  We rewrite the integral as:
\begin{align}
\Uc^{n+1,(k+1)} = \Uc^{n,(k+1)} +
   &\int_{t^n}^{t^{n+1}} \left \{ \left [ \Advs{\Uc}^{(k+1)} - \Advs{\Uc}^{(k)} \right ]
   + \left [ \Rbs{\Uc}^{(k+1)} - \Rbs{\Uc}^{(k)} \right ] \right \} dt + \nonumber \\
  & \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc}^{(k)} + \Rbs{\Uc}^{(k)} \right ] dt
\end{align}
For the first integral, we approximate these terms using a first order
update.  For advection, we simply use the value at the starting time
(analogous to a forward Euler approximation) and for the reactions, we
treat them implicitly by using the value at the ending time (analogous
to a backward Euler approximation).  To gain temporal accuracy, we
further subdivide the time into $M+1$ nodes, and perform this update
over all time nodes for each iteration.  Finally, in \castro\ we use a finite-volume
discretion, so we'll write the state as $\avg{\Uc}$ to indicate that it is an average over the zone.
Denoting the time nodes with
the index $m$, the updated from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
where $\delta t_m = t^{m+1} - t^m$ is the time difference between nodes.
Here, $\avg{\Advs{\Uc}}^m$ is the
average advective term in a zone at time $t^m$, and
$\avg{\Rbs{\Uc}}^{m+1}$ the average of the reactive source at time
$t^{m+1}$.  The second superscript on each term, $(k)$ or $(k+1)$, is
the iteration counter.  This update is an implicit equation for the
new state, $\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is
an integral over the sources constructed from the previous iteration's
values at each time node.  Each iteration of the SDC method increases
the temporal order of accuracy by one, up to the order of accuracy
with which the integral is constructed.

@@

Both the new SDC method and GPU offloading are important to our
science goals, in particular the problem of modeling a flame spreading
across the neutron star during an X-ray burst (XRB).  We discussed the
computational challenges of modeling these events in
\cite{astronum:2018}.  In short: the largest scale we need to model is
the size of the neutron star, or at least several times the scale at
which rotation and the lateral pressure gradient balance (the Rossby
length).  The smallest scale we need to model is dictated by
accurately capturing the burning---we are interested in both the
energy release and nucleosynthesis.  Our first set of two-dimensional simulations
(\cite{astronum:2018,eiden:2019} used two physical approximations to
make the problem more tractable.  First, we used a higher rotation
rate than expected in order to reduce the Rossby length, allowing us
to model a smaller region of that star.  Second, we artificially
boosted the speed of the flame to reduce the duration we need to
model.  Our next set of calculations will relax these approximations.

>>>>

In the SDC formalism~\cite{dutt:2000,minion:2003}, we start by writing our system in integral form:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc} + \Rbs{\Uc} \right ] dt
\end{equation}
We next introduce the idea of iteratively approximating the solution.
We'll use the superscript $k$ to denote the current iteration and
$k+1$ for the new iteration.  We rewrite the integral as:
\begin{align}
\Uc^{n+1,(k+1)} = \Uc^{n,(k+1)} +
   &\int_{t^n}^{t^{n+1}} \left \{ \left [ \Advs{\Uc}^{(k+1)} - \Advs{\Uc}^{(k)} \right ]
   + \left [ \Rbs{\Uc}^{(k+1)} - \Rbs{\Uc}^{(k)} \right ] \right \} dt + \nonumber \\
  & \int_{t^n}^{t^{n+1}} \left [\Advs{\Uc}^{(k)} + \Rbs{\Uc}^{(k)} \right ] dt
\end{align}
For the first integral, we approximate these terms using a first order
update.  For advection, we simply use the value at the starting time
(analogous to a forward Euler approximation) and for the reactions, we
treat them implicitly by using the value at the ending time (analogous
to a backward Euler approximation).  To gain temporal accuracy, we
further subdivide the time into $M+1$ nodes, and perform this update
over all time nodes for each iteration.  Finally, in \castro\ we use a finite-volume
discretization, so we'll write the state as $\avg{\Uc}$ to indicate that it is an average over the zone.
Denoting the time nodes with
the index $m$, the updated from one time node $m$ to $m+1$ appears as:
    \begin{align}
      \label{eq:sdc:general}
      \avg{\Uc}^{{m+1},(k+1)} = \avg{\Uc}^{m,(k+1)}
            &+ \delta t_m \left [ \avg{\Advs{\Uc}}^{m,(k+1)} - \avg{\Advs{\Uc}}^{m,(k)} \right ] \nonumber \\
            &+ \delta t_m \left [ \avg{\Rbs{\Uc}}^{{m+1},(k+1)} - \avg{\Rbs{\Uc}}^{{m+1},(k)} \right ] \nonumber \\
            &+ \int_{t^m}^{t^{m+1}} dt  \left (\avg{\Advs{\Uc}}^{(k)} + \avg{\Rbs{\Uc}}^{(k)}\right)
   \end{align}
where $\delta t_m = t^{m+1} - t^m$ is the time difference between nodes.
Here, $\avg{\Advs{\Uc}}^m$ is the
average advective term in a zone at time $t^m$, and
$\avg{\Rbs{\Uc}}^{m+1}$ the average of the reactive source at time
$t^{m+1}$.  The second superscript on each term, $(k)$ or $(k+1)$, is
the iteration counter.  This update is an implicit equation for the
new state, $\avg{\Uc}^{{m+1},(k+1)}$.  The last term in the update is
an integral over the sources constructed from the previous iteration's
values at each time node.  Each iteration of the SDC method increases
the temporal order of accuracy by one, up to the order of accuracy
with which the integral is constructed.

@@

Both the new SDC method and GPU offloading are important to our
science goals, in particular the problem of modeling a flame spreading
across the neutron star during an X-ray burst (XRB).  We discussed the
computational challenges of modeling these events in
\cite{astronum:2018}.  In short: the largest scale we need to model is
the size of the neutron star, or at least several times the scale at
which rotation and the lateral pressure gradient balance (the Rossby
length).  The smallest scale we need to model is dictated by
accurately capturing the burning---we are interested in both the
energy release and nucleosynthesis.  Our first set of two-dimensional simulations
(\cite{astronum:2018,eiden:2019} used two physical approximations to
make the problem more tractable.  First, we used a higher rotation
rate than expected in order to reduce the Rossby length, allowing us
to model a smaller region of that star.  Second, we artificially
boosted the speed of the flame to reduce the duration we need to
model.  Figure~\ref{fig:xrb} shows the structure of the flame.  Our next set of calculations will relax these approximations.
